
# ==============================================================================
# SCRIPT: Impact of Rice Variety and Inoculation on Root Microbiota (16S)
# AUTHOR: Léa Jobert
# DATE: February 2026
# ==============================================================================


#1-Data preparation
##1.1-Install packages
```{r message=FALSE, warning=FALSE}
library(dada2)
library(phyloseq)
library(microViz)
library(dplyr)
library(data.table)
library(ggplot2)
library(vegan)
library(phangorn)
library(picante)
library(tidyverse)
library(patchwork)
library(vctrs)
library(microbiomeutilities)
library(pairwiseAdonis)
library(pheatmap)
library(RColorBrewer)
library(decontam)
library(ranacapa)
library(multcompView)
library(scales)
library(DECIPHER)
library(devtools)
library(ggVennDiagram)
library(openxlsx)
library(DESeq2)
library(ALDEx2)
library(ggrepel)
library(ggpubr)
library(ggVennDiagram)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(forcats)
library(RColorBrewer)
```



# 2. FILE PATHS & DIRECTORIES 
```{r}
# Set working directory - # Change to your directory
base_path <- "path"
setwd(base_path)

# Create output directory for figures
output_dir_fig <- file.path(base_path, "output_path")
if(!dir.exists(output_dir_fig)) dir.create(output_dir_fig)
```



# 3. DATA IMPORT & PHYLOSEQ CREATION 
```{r}
# Import raw ASV and Taxonomy data from CSV
raw_data_path <- file.path(base_path, "input.csv")
df_raw <- read.csv(raw_data_path, sep=";", header=TRUE, stringsAsFactors=FALSE)

# Create Taxonomy Table
tax_table_df <- df_raw[, c("ASV", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")]
rownames(tax_table_df) <- tax_table_df$ASV
tax_table_df$ASV <- NULL

# Create ASV Table
# Removing taxonomy columns to keep only samples
asv_table <- df_raw[, setdiff(names(df_raw), c("taxonomy", "ASV", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"))]
rownames(asv_table) <- asv_table$ID
asv_table$ID <- NULL
asv_table <- t(asv_table)
colnames(asv_table) <- df_raw$ASV

# Import Metadata
metadata_path <- file.path(base_path, "Metadata_Varieties.csv")
metadata <- read.csv2(metadata_path)
rownames(metadata) <- metadata$SampleID

# Ensure consistency between ASV table and Metadata
common_samples <- intersect(rownames(asv_table), rownames(metadata))
asv_table <- asv_table[common_samples, ]
metadata <- metadata[common_samples, ]

# Convert ASV table to numeric matrix
asv_matrix <- apply(asv_table, 2, as.numeric)
rownames(asv_matrix) <- rownames(asv_table)

# Create Phyloseq Object
ps16S_raw <- phyloseq(otu_table(asv_matrix, taxa_are_rows=FALSE), 
                  sample_data(metadata), 
                  tax_table(as.matrix(tax_table_df)))
```


# 4. DATA FILTERING 
```{r}

# Remove non-bacterial/archaeal sequences (Plants/Viridiplantae)
ps16S_filtered <- ps16S_raw %>% subset_taxa(Kingdom != "k__Viridiplantae")

# Remove NA at Kingdom and Phylum levels
ps16S_filtered <- subset_taxa(ps16S_filtered, !is.na(Kingdom) & !is.na(Phylum))

# Remove Mitochondria and Chloroplast contamination
ps16S_filtered <- ps16S_filtered %>% 
  subset_taxa(Family != "Mitochondria" & Class != "Chloroplast")

# Final cleaning for the root dataset (excluding specific varieties or outliers if needed)
# Example: removing "PhkarRumduol_GM" as requested
ps16S_Root <- subset_samples(ps16S_filtered, Variety != "PhkarRumduol_GM")
# Remove a specific ASV by its name
ps16S_Root <- prune_taxa(taxa_names(ps16S_Root) != "16125", ps16S_Root)

# Remove low-abundance taxa (Total count < 10)
ps16S_Root_10 <- phyloseq::filter_taxa(ps16S_Root, function(x) sum(x) >= 10, prune = TRUE)
```


# 5. NORMALIZATION & SUBSETTING 
```{r}
# Rarefaction to minimum depth for diversity analyses
ps16S_Root_10_rare <- rarefy_even_depth(ps16S_Root_10, rngseed = 123, 
                                 sample.size = min(sample_sums(ps16S_Root_10)), 
                                 replace = FALSE, verbose = TRUE)

# Control group subsets
ps16S_control <- subset_samples(ps16S_Root, Condition == "Control")
ps16S_control_10 <- filter_taxa(ps16S_control, function(x) sum(x) >= 10, prune = TRUE)
ps16S_control_10_rare <- rarefy_even_depth(ps16S_control_10, rngseed = 123, 
                                 sample.size = min(sample_sums(ps16S_control_10)), replace = FALSE)

# Inoculated group subsets
ps16S_inoc <- subset_samples(ps16S_Root, Condition == "Inoculated")
ps16S_inoc_10 <- filter_taxa(ps16S_inoc, function(x) sum(x) >= 10, prune = TRUE)
ps16S_inoc_10_rare <- rarefy_even_depth(ps16S_inoc_10, rngseed = 123, 
                                  sample.size = min(sample_sums(ps16S_inoc_10)), replace = FALSE)
```



#6. SYMPTOMS IMPORT
```{r}
# ==============================================================================
# 6. SYMPTOMS & SIZE IMPORT (CLEAN MERGE)
# ==============================================================================

# 1. Import CSV
# Assumes the file is in your working directory set in Section 2
symp_df <- read.csv("Symptoms.csv", sep = ";", header = TRUE, stringsAsFactors = FALSE)

# 2. Clean and Format Data
symp_df <- symp_df %>%
  filter(!is.na(SampleID) & SampleID != "") %>%
  mutate(
    Symptoms_up = as.numeric(Symptoms_up),
    Size        = as.numeric(Size)         # Added Size column
  ) %>%
  # Select only relevant columns to avoid duplicates/confusion
  dplyr::select(SampleID, Symptoms_up, Size)

# 3. Merge into Phyloseq Object
# First, extract current metadata from the rarefied object
curr_meta <- data.frame(sample_data(ps16S_Root_10_rare))
curr_meta$SampleID <- rownames(curr_meta)

# Remove any existing Symptoms/Size columns to prevent duplicates (.x, .y)
# We keep only the core experimental design columns
core_cols <- c("SampleID", "Variety", "Condition", "Group")
curr_meta_clean <- curr_meta[, colnames(curr_meta) %in% core_cols]

# Join with new data
new_metadata <- left_join(curr_meta_clean, symp_df, by = "SampleID")
rownames(new_metadata) <- new_metadata$SampleID

# 4. UPDATE THE PHYLOSEQ OBJECT (Crucial Step)
# This ensures Fig 4B and the Correlation analysis can access the data
sample_data(ps16S_Root_10_rare) <- sample_data(new_metadata)

# Verification
message("Columns in Phyloseq Metadata:")
print(colnames(sample_data(ps16S_Root_10_rare)))
# Should show: SampleID, Variety, Condition, Group, Symptoms_up, Size
```


#7. SETTING COLORS PALET
```{r}

colors_Variety <- c( "Leak" = "#7CAE00", "PhkarRumduol" = "#00BFC4", "PhkarMalis"="#FF3B0A","Russey"="#FFC000", "IR64"="#9E1FFF" )
group_colors <- c(
                "IR64_Control" = "#9E1FFF", "IR64_Inoculated" = "#CA85FF", 
                "Leak_Control" = "#7CAE00", "Leak_Inoculated" = "#AFF500", 
                "PhkarRumduol_Control" = "#00BFC4", "PhkarRumduol_Inoculated" = "#00EDF5", 
                "PhkarMalis_Control"="#FF3B0A","PhkarMalis_Inoculated"="#FF7C5C",
                "Russey_Control"="#FFC000","Russey_Inoculated"="#FFD65C")

condition_colors <- c("Control" = "#009E73", "Inoculated" = "#D55E00")


#palettes pour taxonomic binning
colors_taxa <- c(
  "#1b9e77", "#d95f02", "#38369A", "#e7298a", "#66a61e",
  "#e6ab02", "#FF3200FF", "#666666", "#a6cee3", "#CA562CFF",
  "#b2df8a", "#33a02c", "#C490D1", "#e31a1c", "#fdbf6f",
  "#94B0DA", "#009392FF", "#B2182BFF", "#99CE64FF",
  "#8dd3c7", "#F8D564FF", "#fb9a99", "#fb8072", "#80b1d3",
  "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd",
  "gray70"  # pour le groupe "Autres"
)

#pour les Venn: 
color_venn  <- c(low = "#F2EBBBFF", high = "#7F1019")


#🦠 🐛 🐜 🐝 🌱 🌿 🍃 🌾 🍄 📊 📈 📉 🗂️ 📑🎯 📌 🏷️ 📦 🗃🚀 ✨ 🔍 🧠 🔄 💡 🦠 🧪 📊 🚀✅ 

```


# 8. ANALYSIS & FIGURES --------------------------------------------------------
# --- FIG 1: TAXONOMIC HEATMAP (Top 25 Species) ---
```{r}
# ==============================================================================
# FIG 1: TAXONOMIC HEATMAP (Top 30 Species - Roots)
# ==============================================================================
library(phyloseq)
library(dplyr)
library(tidyr)
library(openxlsx)
library(ComplexHeatmap)
library(circlize)
library(tibble)

# --- 1. Data Preparation & Name Cleaning ---

# Filter Control (No IR64)
ps16S_control_sub <- subset_samples(ps16S_control, Variety != "IR64")

# Agglomerate to Species
ps_species <- tax_glom(ps16S_control_sub, taxrank = "Species", NArm = FALSE)

# Relative Abundance
ps_species_rel <- transform_sample_counts(ps_species, function(x) 100 * x / sum(x))

# Melt
df_melted <- psmelt(ps_species_rel)

# Clean Names (Space separator + "sp." for NAs)
df_melted <- df_melted %>%
  mutate(
    Species_Clean = ifelse(is.na(Species) | Species == "NA" | Species == "", "sp.", Species),
    Genus_Species = paste(Genus, Species_Clean, sep = " ")
  )

# --- Data Aggregation and Global Ranking ---

# 1. Calculate mean relative abundance per variety for each species
mean_abundance_per_variety <- df_melted %>%
  group_by(Variety, Genus_Species) %>%
  summarise(MeanAbundance = mean(Abundance), .groups = "drop")

# 2. Identify the Top 30 species based on cumulative abundance across all varieties
# (Sum of means across the 4 varieties)
top30_global_list <- mean_abundance_per_variety %>%
  group_by(Genus_Species) %>%
  summarise(TotalAbundance = sum(MeanAbundance)) %>%
  arrange(desc(TotalAbundance)) %>%  # Sort highest to lowest
  slice_head(n = 30)                 # Select top 30

# 3. Filter dataset to keep only the Top 30 and pivot to wide format
top30_wide <- mean_abundance_per_variety %>%
  filter(Genus_Species %in% top30_global_list$Genus_Species) %>%
  pivot_wider(names_from = Variety, values_from = MeanAbundance, values_fill = 0) %>%
  # Crucial: Reorder rows to match the global ranking (Descending Total Abundance)
  arrange(match(Genus_Species, top30_global_list$Genus_Species))

# Save Excel
excel_file <- file.path(output_dir_fig, "Table_Top30_Species_by_Variety_Root.xlsx")
write.xlsx(top30_wide, file = excel_file)
message("✅ Excel file saved: ", excel_file)

# --- 2. Heatmap Setup ---

# Convert to Matrix
mat_root <- top30_wide %>%
  column_to_rownames("Genus_Species") %>%
  as.matrix()

# Annotations (Varieties)
annotation_df <- data.frame(Variety = colnames(mat_root))
rownames(annotation_df) <- colnames(mat_root)

# Ensure colors_Variety is defined in your environment
ann_colors <- list(Variety = colors_Variety)

# --- 3. Colors (Root Specific Palette) ---
col_fun_root <- colorRamp2(
  c(0, 0.2, 0.5, 1, 2, 3, 4, 5, 7, 10, 20),
  c("white", "#EDF7DE", "#E4F2CF", "#D2E9AF", "#B7DC7F", "#93CA3F", "#7DB030", "#5B8023", "#44601A", "#22300D", "black")
)

# --- 4. Generate Heatmap ---
heatmap_root <- Heatmap(
  mat_root,
  name = "Abundance (%)",
  
  # Clustering
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_order = colnames(mat_root),
  
  # Hide Tree
  show_row_dend = FALSE,
  
  # Cell Size (Square look)
  width = unit(5, "cm"),
  
  # Annotation
  top_annotation = HeatmapAnnotation(df = annotation_df, col = ann_colors),
  
  # Colors
  col = col_fun_root,
  
  # Typography (Italic for Species)
  row_names_side = "right",
  row_names_gp = gpar(fontsize = 10, fontface = "italic"),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  
  # Borders
  rect_gp = gpar(col = "white", lwd = 1)
)

# --- 5. Save Figure ---
svg_file <- file.path(output_dir_fig, "Fig1_Heatmap_Top30_Root_Final.svg")

svg(svg_file, width = 7, height = 8)
draw(heatmap_root)
dev.off()

message("✅ Final Root Heatmap saved to: ", svg_file)
```
# --- FIG 2A: ALPHA DIVERSITY (Richness) ---
```{r}
alpha_div <- estimate_richness(ps16S_control_10, measures = c("Observed", "Shannon"))
alpha_data <- cbind(as.data.frame(sample_data(ps16S_control_10)), alpha_div)

# Kruskal-Wallis Test
kw_test <- kruskal.test(Observed ~ Variety, data = alpha_data)

p_alpha <- ggplot(alpha_data, aes(x = Variety, y = Observed, fill = Variety)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(color = "black", size = 2, alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = colors_Variety) +
  labs(title = "A. Observed ASV Richness (Control)", y = "Number of ASVs") +
  annotate("text", x = 1, y = 800, 
           label = paste("Kruskal-Wallis p =", signif(kw_test$p.value, 3)), hjust = 0) +
  theme_minimal() + theme(axis.title.x = element_blank())

ggsave(file.path(output_dir_fig, "Fig2A_Alpha_Diversity.svg"), p_alpha, width = 6, height = 4)
```

# --- FIG 2C: BETA DIVERSITY (NMDS - Control plants) ---
```{r}
# 1. Calculate Bray-Curtis distance and NMDS ordination
bray_dist_ctrl <- phyloseq::distance(ps16S_control_10_rare, method = "bray")
nmds_ctrl <- ordinate(ps16S_control_10_rare, method = "NMDS", distance = "bray")

# 2. Run PERMANOVA to get R2 and p-value
# Ensure metadata is formatted as a dataframe
ctrl_metadata <- as.data.frame(as.matrix(sample_data(ps16S_control_10_rare)))
adonis_ctrl <- adonis2(bray_dist_ctrl ~ Variety, data = ctrl_metadata)

# 3. Extract statistical values for the subtitle
r2_val     <- round(adonis_ctrl$R2[1], 3)
p_val      <- adonis_ctrl$`Pr(>F)`[1]
stress_val <- round(nmds_ctrl$stress, 3) # Extracting stress value

# 4. Generate the Plot
p_nmds <- plot_ordination(ps16S_control_10_rare, nmds_ctrl, color = "Variety") +
  geom_point(size = 4, alpha = 0.7) +
  stat_ellipse(aes(fill = Variety), geom = "polygon", alpha = 0.1, color = NA) +
  scale_color_manual(values = colors_Variety) +
  scale_fill_manual(values = colors_Variety) +
  labs(
    title = "C. NMDS - Bacterial Community (Control Roots)", 
    subtitle = paste0(" | R² = ", r2_val, " | p = ", p_val, " | Stress = ", stress_val)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(face = "italic", size = 11),
    legend.title = element_text(face = "bold")
  )

# Display plot
print(p_nmds)

# Save figure

ggsave(file.path(output_dir_fig, "Fig2C_NMDS_Control.svg"), p_nmds, width = 5, height = 3)
```

# --- FIG 2E: COMPLEMENTARY TO 2C ---
```{r}
# ==============================================================================
# FIG 2E: Complementary analyses to NMDS (Distances, Dispersion, Pairwise tests)
# ==============================================================================

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(vegan)
library(patchwork)

set.seed(123)  # reproducibility

# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# 1) Sample metadata
# ------------------------------------------------------------------------------

meta_ctrl <- data.frame(sample_data(ps16S_control_10_rare))

meta_ctrl$SampleID <- rownames(meta_ctrl)

meta_ctrl <- meta_ctrl %>%
  dplyr::select(SampleID, Variety)

# Ensure factor order matches color palette
lvl <- intersect(names(colors_Variety), unique(meta_ctrl$Variety))
meta_ctrl$Variety <- factor(meta_ctrl$Variety, levels = lvl)

# ------------------------------------------------------------------------------
# 2) Bray–Curtis distance matrix
# ------------------------------------------------------------------------------
otu_mat <- as.matrix(otu_table(ps16S_control_10_rare))
bray_dist <- vegdist(otu_mat, method = "bray")
bray_mat  <- as.matrix(bray_dist)

# ------------------------------------------------------------------------------
# 3) Pairwise sample distances
# ------------------------------------------------------------------------------
pairs_df <- as.data.frame(t(combn(rownames(bray_mat), 2)), stringsAsFactors = FALSE)
colnames(pairs_df) <- c("Sample1", "Sample2")

pairs_df$Distance <- bray_mat[
  cbind(match(pairs_df$Sample1, rownames(bray_mat)),
        match(pairs_df$Sample2, colnames(bray_mat)))
]

pairs_df <- pairs_df %>%
  dplyr::left_join(
    meta_ctrl %>% dplyr::select(SampleID, Variety),
    by = c("Sample1" = "SampleID")
  ) %>%
  dplyr::rename(Var1 = Variety) %>%
  dplyr::left_join(
    meta_ctrl %>% dplyr::select(SampleID, Variety),
    by = c("Sample2" = "SampleID")
  ) %>%
  dplyr::rename(Var2 = Variety) %>%
  dplyr::mutate(
    Comparison = ifelse(Var1 == Var2, "Intra-variety", "Inter-variety"),
    Pair = paste(
      pmin(as.character(Var1), as.character(Var2)),
      pmax(as.character(Var1), as.character(Var2)),
      sep = " vs "
    )
  )


# ------------------------------------------------------------------------------
# 4) Intra vs Inter distances
# ------------------------------------------------------------------------------
p_intra_inter <- ggplot(pairs_df, aes(x = Comparison, y = Distance)) +
  geom_violin(trim = FALSE, fill = "grey90") +
  geom_boxplot(width = 0.25, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.15, size = 0.6) +
  labs(
    title = "Community dissimilarity: intra vs inter-variety",
    y = "Bray–Curtis distance",
    x = NULL,
    subtitle = paste(
      "Wilcoxon p =",
      signif(wilcox.test(Distance ~ Comparison, data = pairs_df)$p.value, 3)
    )
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

# ------------------------------------------------------------------------------
# 5) Multivariate dispersion (betadisper)
# ------------------------------------------------------------------------------
bd <- betadisper(bray_dist, meta_ctrl$Variety)

disp_df <- data.frame(
  SampleID = names(bd$distances),
  Distance = bd$distances,
  Variety  = bd$group
)

p_disp <- ggplot(disp_df, aes(x = Variety, y = Distance, fill = Variety)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = colors_Variety) +
  labs(
    title = "Multivariate dispersion (betadisper)",
    subtitle = "Distance to group centroid",
    y = "Distance to centroid",
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(face = "bold")
  )

# ------------------------------------------------------------------------------
# 6) Pairwise PERMANOVA (manual)
# ------------------------------------------------------------------------------
pairs_levels <- combn(levels(meta_ctrl$Variety), 2, simplify = FALSE)

pw_table <- lapply(pairs_levels, function(p) {

  sub_meta <- meta_ctrl %>% filter(Variety %in% p)
  sub_dist <- as.dist(bray_mat[sub_meta$SampleID, sub_meta$SampleID])

  ad <- adonis2(sub_dist ~ Variety, data = sub_meta, permutations = 999)

  data.frame(
    Var1 = p[1],
    Var2 = p[2],
    R2   = ad$R2[1],
    pval = ad$`Pr(>F)`[1]
  )
}) %>%
  bind_rows() %>%
  mutate(p_adj = p.adjust(pval, method = "BH"))

# ------------------------------------------------------------------------------
# 7) Heatmap: mean distances + PERMANOVA significance
# ------------------------------------------------------------------------------
mean_dist <- pairs_df %>%
  group_by(Var1, Var2) %>%
  summarise(MeanDist = mean(Distance), .groups = "drop")

heatmap_df <- expand.grid(Var1 = lvl, Var2 = lvl) %>%
  left_join(mean_dist, by = c("Var1", "Var2")) %>%
  left_join(mean_dist, by = c("Var1" = "Var2", "Var2" = "Var1"),
            suffix = c("", "_rev")) %>%
  mutate(MeanDist = coalesce(MeanDist, MeanDist_rev))

stars_df <- pw_table %>%
  mutate(
    Label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    ),
    Var1 = factor(Var1, levels = lvl),
    Var2 = factor(Var2, levels = lvl)
  )

p_heat <- ggplot(heatmap_df, aes(x = Var1, y = Var2, fill = MeanDist)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "YlOrRd", name = "Mean Bray–Curtis\nDistance") +
geom_text(
  data = stars_df,
  aes(x = Var1, y = Var2, label = Label),
  inherit.aes = FALSE,
  size = 6
)+
  labs(
    title = "Pairwise community dissimilarity and PERMANOVA",
    subtitle = "Color = mean Bray–Curtis distance | Stars = FDR-adjusted p-value",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ------------------------------------------------------------------------------
# 8) Combine panels (optional)
# ------------------------------------------------------------------------------
(p_intra_inter | p_disp) / p_heat

```





# --- FIG 3: VENN DIAGRAM & CORE MICROBIOTA ---
```{r}
# ==============================================================================
# 1. DATA PREPARATION AND FILTERING FOR VENN DIAGRAM
# ==============================================================================

# Using the filtered phyloseq object (Control samples, 10 reads filter)
# Extracting tables (maintaining original row names)
otu_controls <- as.data.frame(otu_table(ps16S_control_10))
metadata_controls <- as.data.frame(sample_data(ps16S_control_10))

# Core filtering parameters per variety
min_abundance <- 100     # Minimum total reads per variety
min_prev      <- 0.8     # Minimum prevalence (80% of samples)

# Identifying varieties present in the dataset
varieties <- unique(metadata_controls$Variety)

# Extracting ASVs per variety based on filters
otu_listF <- lapply(varieties, function(var) {
  # Subset samples for the specific variety
  samples <- rownames(metadata_controls[metadata_controls$Variety == var, ])
  samples <- intersect(samples, rownames(otu_controls))
  
  if (length(samples) == 0) return(character(0))
  
  # Variety-specific OTU subset
  otu_subset <- otu_controls[samples, , drop = FALSE]
  
  # Filter 1: Total abundance within this variety >= min_abundance
  pass_abundance <- colSums(otu_subset) >= min_abundance
  
  # Filter 2: Prevalence (present in >= 80% of samples)
  pass_prevalence <- (colSums(otu_subset > 0) / nrow(otu_subset)) >= min_prev
  
  # Final selection: ASVs passing both filters
  keep_asv <- pass_abundance & pass_prevalence
  
  return(colnames(otu_subset)[keep_asv])
})

names(otu_listF) <- varieties
otu_list_venn_ct <- otu_listF


# ==============================================================================
# 2. CORE MICROBIOTA STATISTICS
# ==============================================================================

# Define the "Core Microbiota" as the intersection of all varieties
core_asvs <- Reduce(intersect, otu_list_venn_ct)

# Global metrics for the dataset (ps16S_control_10)
total_asvs_count <- ntaxa(ps16S_control_10_rare)                   
total_reads_count <- sum(sample_sums(ps16S_control_10_rare))       

# Specific metrics for the Core
core_asvs_count <- length(core_asvs)                 
core_reads_count <- sum(otu_controls[, colnames(otu_controls) %in% core_asvs])

# Proportion calculations
prop_diversity <- (core_asvs_count / total_asvs_count) * 100
prop_abundance <- (core_reads_count / total_reads_count) * 100

# Print statistics to console
cat("\n=================================================\n")
cat("📊 CORE MICROBIOTA STATISTICS\n")
cat("=================================================\n")
cat("Total ASVs in study           : ", total_asvs_count, "\n")
cat("Core ASVs count               : ", core_asvs_count, "\n")
cat("👉 Proportion of Diversity    : ", round(prop_diversity, 2), "%\n\n")

cat("Total Reads in dataset        : ", total_reads_count, "\n")
cat("Core Reads count              : ", core_reads_count, "\n")
cat("👉 Proportion of Abundance    : ", round(prop_abundance, 2), "%\n")
cat("=================================================\n\n")

# ==============================================================================
# 3. VENN DIAGRAM VISUALIZATION
# ==============================================================================

# Color palette definition
colors_Variety <- c(
  "Leak"         = "#7CAE00", 
  "PhkarRumduol" = "#00BFC4", 
  "PhkarMalis"   = "#FF3B0A",
  "Russey"       = "#FFC000", 
  "IR64"         = "#9E1FFF"
)

# Align colors with the order of varieties in the list
colors_ordered <- colors_Variety[names(otu_list_venn_ct)]

# Generate Venn Diagram
venn <- ggVennDiagram(otu_list_venn_ct, 
                      set_color = colors_ordered,    
                      set_size = 5,                  
                      label = "count",               
                      label_alpha = 0) +             
  scale_fill_gradient(low = "#F2EBBBFF", high = "#7F1019", name = "ASV Count") +
  labs(title = "Core Microbiota - Shared ASVs (Control Roots)") +
  theme(legend.position = "right")

print(venn)

# Save figure - # Change to your directory
ggsave(filename = file.path(output_dir_fig, "Fig3_Venn_Core_Roots.svg"), 
       plot = venn, width = 8, height = 6)




# ==============================================================================
# 4. DONUT CHART (Core Taxonomic Composition - Genus)
# ==============================================================================

rank_to_show <- "Genus"
min_pct_other <- 2 

tax_df <- as.data.frame(tax_table(ps16S_control_10_rare))
tax_df$ASV <- rownames(tax_df)

# Subset OTU table to include only Core ASVs
otu_core_subset <- otu_controls[, colnames(otu_controls) %in% core_asvs, drop=FALSE]
core_abundance_vec <- colSums(otu_core_subset)

core_tax_df <- data.frame(
  ASV = names(core_abundance_vec),
  Abundance = as.numeric(core_abundance_vec)
) %>%
  left_join(tax_df, by = "ASV")

# Summarize core composition by Genus
core_summary <- core_tax_df %>%
  filter(!is.na(.data[[rank_to_show]]), 
         .data[[rank_to_show]] != "Unclassified") %>% 
  group_by(.data[[rank_to_show]]) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  mutate(pct = 100 * TotalAbundance / sum(TotalAbundance)) %>%
  
  # Group low abundance genera into "Other"
  mutate(DisplayGroup = ifelse(pct < min_pct_other, "Other", .data[[rank_to_show]])) %>%
  group_by(DisplayGroup) %>%
  summarise(TotalAbundance = sum(TotalAbundance), .groups = "drop") %>%
  mutate(pct = round(100 * TotalAbundance / sum(TotalAbundance), 1)) %>%
  arrange(desc(pct))

# Set colors
lvls <- core_summary$DisplayGroup
pal_vec <- colorRampPalette(brewer.pal(12, "Set3"))(length(lvls))
names(pal_vec) <- lvls
if("Other" %in% names(pal_vec)) pal_vec["Other"] <- "grey80"

# Plot Genus Donut Chart
donut_core <- ggplot(core_summary, aes(x = 2, y = pct, fill = fct_reorder(DisplayGroup, pct))) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void() +
  geom_text(aes(label = paste0(DisplayGroup, " (", pct, "%)")),
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_manual(values = pal_vec, name = rank_to_show) +
  labs(title = paste0("Core Microbiota Composition (", rank_to_show, ")"),
       subtitle = paste0("Represents ", round(prop_abundance, 1), "% of total reads in dataset"))

print(donut_core)

ggsave(filename = file.path(output_dir_fig, "Fig3_Core_Donut_Genus.svg"), 
       plot = donut_core, width = 8, height = 6)


# ==============================================================================
# 5. DONUT CHART (Core Taxonomic Composition - Family)
# ==============================================================================

rank_to_show <- "Genus"
rank_parent <- "Family"
min_pct_other <- 2       
min_pct_label <- 2       

tax_df <- as.data.frame(tax_table(ps16S_control_10_rare))
tax_df$ASV <- rownames(tax_df)

otu_mat <- as.data.frame(otu_table(ps16S_control_10_rare))
otu_core <- otu_mat[, colnames(otu_mat) %in% core_asvs, drop = FALSE]
core_abundance <- colSums(otu_core)

df_core <- data.frame(ASV = names(core_abundance), Abundance = core_abundance) %>%
  left_join(tax_df, by = "ASV")

# Aggregation logic
core_summary_fam <- df_core %>%
  filter(!is.na(.data[[rank_to_show]]), 
         .data[[rank_to_show]] != "Unclassified") %>%
  group_by(.data[[rank_to_show]], .data[[rank_parent]]) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  mutate(pct = 100 * Abundance / sum(Abundance)) %>%
  mutate(
    Genus_Group = ifelse(pct < min_pct_other, "Other", .data[[rank_to_show]]),
    Family_Group = ifelse(pct < min_pct_other, "Other", .data[[rank_parent]])
  ) %>%
  group_by(Genus_Group, Family_Group) %>%
  summarise(pct = sum(pct), .groups = "drop") %>%
  arrange(Family_Group, desc(pct))

# Reorder factors for consistent plotting
core_summary_fam$Genus_Group <- factor(core_summary_fam$Genus_Group, levels = core_summary_fam$Genus_Group)
levels_fam <- unique(core_summary_fam$Family_Group)
levels_fam <- c(setdiff(levels_fam, "Other"), "Other")
core_summary_fam$Family_Group <- factor(core_summary_fam$Family_Group, levels = levels_fam)

core_summary_fam <- core_summary_fam %>%
  mutate(Label = ifelse(pct >= min_pct_label, paste0(Genus_Group, "\n(", round(pct, 1), "%)"), ""))

# Defined Family color palette
family_colors <- c(
  "Burkholderiaceae"      = "#FC7A57",  
  "Comamonadaceae"        = "#28666E",  
  "Rhizobiaceae"          = "#B5B682",  
  "Xanthobacteraceae"     = "#B2DF8A",  
  "Sphingomonadaceae"     = "#FCD757",  
  "Beijerinckiaceae"      = "#FDBF6F",  
  "Chitinophagaceae"      = "#574D68",  
  "Solirubrobacteraceae"  = "#CAB2D6",  
  "Solimonadaceae"        = "#94D1BE",  
  "Holophagaceae"         = "#B15928",  
  "Other"                 = "#D3D3D3"   
)

# Plot Family Donut Chart
donut_core_fam <- ggplot(core_summary_fam, aes(x = 2, y = pct, fill = Family_Group)) +
  geom_col(width = 1, color = "white", linewidth = 1) + 
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5),
            size = 3.5, fontface = "bold", color = "black") + 
  scale_fill_manual(values = family_colors, name = "Family") +
  theme_void() +
  labs(title = paste0("Core Microbiota (", rank_to_show, ")"),
       subtitle = "Grouped by Family") +
  theme(legend.position = "right", legend.title = element_text(face="bold"))

print(donut_core_fam)

ggsave(file.path(output_dir_fig, "Fig3_Core_Donut_Family_Custom.svg"),
  donut_core_fam, width = 10, height = 7)
```


# --- FIG 4.A: ALPHA, IMPACT OF INOCULATION ---
```{r}

library(ggplot2)
library(ggpubr)

#Table of alpha-diversity estimators
table_variety <- estimate_richness(ps16S_Root_10, split = TRUE, measures=c("Observed", "Shannon"))
#Add evenness
H.variety <- table_variety$Shannon
S1.variety <- table_variety$Observed
S.variety <- log(S1.variety)
eve_variety <- H.variety/S.variety
table_variety$Evenness <- eve_variety
#Bind sampledata + table of alpha_div (here R1 only)
datar.variety <- cbind(sample_data(ps16S_Root_10), table_variety)



# Définition des couleurs personnalisées
group_colors <- c(
                "IR64_Control" = "#9E1FFF", "IR64_Inoculated" = "#CA85FF", 
                "Leak_Control" = "#7CAE00", "Leak_Inoculated" = "#AFF500", 
                "PhkarRumduol_Control" = "#00BFC4", "PhkarRumduol_Inoculated" = "#00EDF5", 
                "PhkarMalis_Control"="#FF3B0A","PhkarMalis_Inoculated"="#FF7C5C",
                "Russey_Control"="#FFC000","Russey_Inoculated"="#FFD65C")

datar.variety$Inoculation <- ifelse(
  grepl("Control", datar.variety$Group),
  "Control",
  "Inoculated"
)

# Graphiques avec corrections

p.ric.variety <- ggplot(
  datar.variety,
  aes(x = Group, y = Observed, fill = Group, colour = Inoculation)
) +
  geom_boxplot(outlier.shape = NA, linewidth = 1) +
  geom_jitter(
    aes(color = Inoculation),
    size = 2,
    alpha = 0.6
  ) +  scale_fill_manual(values = group_colors) +
  scale_colour_manual(values = condition_colors) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  ggtitle("A. Observed ASV Richness") +
  stat_compare_means(
    aes(label = ..p.signif..),
    method = "wilcox.test",
    comparisons = list(
      c("IR64_Control", "IR64_Inoculated"),
      c("Leak_Control", "Leak_Inoculated"),
      c("PhkarRumduol_Control", "PhkarRumduol_Inoculated"),
      c("PhkarMalis_Control", "PhkarMalis_Inoculated"),
      c("Russey_Control", "Russey_Inoculated")
    )
  )


p.ric.variety
ggsave(p.ric.variety, output_dir_fig, dpi=300, device = "svg", width = 8, height = 4, filename = "Fig4.A_richness_variety_condition.svg")

```

# --- FIG 4.B: BETA, IMPACT OF INOCULATION ---
```{r}
library(vegan)
library(ggplot2)
library(dplyr)
library(phyloseq)

# ==============================================================================
# 1. DATA PREPARATION
# ==============================================================================
# Re-extract metadata to ensure we have Symptoms and Size columns available
sample_metadata <- data.frame(sample_data(ps16S_Root_10_rare))
sample_metadata$SampleID <- rownames(sample_metadata)

# Calculate Bray-Curtis distance matrix
otu_mat <- as.matrix(otu_table(ps16S_Root_10_rare))
bray_dist <- vegdist(otu_mat, method = "bray")

# Calculate Global Range for Size Legend (Standardization)
global_symptom_range <- range(sample_metadata$Symptoms_up, na.rm = TRUE)

# ==============================================================================
# 2. PLOT GENERATION LOOP
# ==============================================================================
variety_list <- unique(sample_metadata$Variety)

for (current_var in variety_list) {
  
  # A. Subset data
  subset_info <- sample_metadata %>% filter(Variety == current_var)
  
  if (nrow(subset_info) < 3) next
  
  # B. Subset distance
  subset_dist <- as.dist(as.matrix(bray_dist)[subset_info$SampleID, subset_info$SampleID])
  
  # C. Run NMDS
  nmds_res <- tryCatch({
    metaMDS(subset_dist, k = 2, trymax = 100, trace = FALSE)
  }, error = function(e) { return(NULL) })
  
  if (is.null(nmds_res)) next
  
  # D. Prepare Plotting Data
  nmds_pts <- as.data.frame(nmds_res$points)
  nmds_pts$SampleID <- rownames(nmds_pts)
  nmds_plot_df <- left_join(nmds_pts, subset_info, by = "SampleID")
  
  # Group for Fill color (Variety_Condition)
  nmds_plot_df$Group <- paste(current_var, nmds_plot_df$Condition, sep = "_")
  
  # E. Statistics (PERMANOVA)
  perm_res <- adonis2(subset_dist ~ Condition, data = subset_info, permutations = 999)
  
  # F. Visualization
  stats_label <- paste0("Stress = ", round(nmds_res$stress, 3), 
                        " | R² = ", round(perm_res$R2[1], 3), 
                        " | p = ", round(perm_res$`Pr(>F)`[1], 3))
  
  p_var <- ggplot(nmds_plot_df, aes(x = MDS1, y = MDS2, 
                                    color = Condition, # Outline Color (Points + Ellipse)
                                    fill = Group,      # Inside Color (Ellipse only)
                                    size = Symptoms_up)) +
    
    # Ellipses: Removed 'color = NA' so it inherits 'color = Condition'
    stat_ellipse(geom = "polygon", alpha = 0.2) +
    
    geom_point(alpha = 0.8) +
    
    # Colors
    scale_color_manual(values = condition_colors) + 
    scale_fill_manual(values = group_colors) +      
    
    # Size Scale (Standardized)
    scale_size_continuous(
      name = "Symptoms (cm)", 
      range = c(2, 8),               
      limits = global_symptom_range,
      breaks = waiver()
    ) +
    
    # LEGEND MANAGEMENT: Remove Fill (Group), Keep Color (Condition) & Size
    guides(fill = "none") +
    
    theme_minimal(base_size = 14) +
    labs(title = paste("Variety:", current_var), subtitle = stats_label) +
    theme(plot.subtitle = element_text(size = 14, face = "italic"))

  # G. Save Figure
  fig_name <- paste0("Fig4_NMDS_", current_var, "_Inoculation_Impact.svg")
  ggsave(filename = file.path(output_dir_fig, fig_name), plot = p_var, width = 6, height = 5)
}

message("✅ Updated Figure 4 (Colored outlines, cleaned legends) saved.")
```


# --- FIG 5: GLOBAL EFFECT OF INOCULATION (DESEQ_MANHATTAN PLOT) ---
```{r}
library(phyloseq)
library(DESeq2)
library(dplyr)
library(apeglm)
library(ggplot2)

# 1) Data Input: Use the NON-rarefied phyloseq object
ps <- ps16S_Root_10  

# 2) DESeq2 Processing -----------------------------------------------------
# Model design: account for Variety background to isolate the Inoculation effect
dds <- phyloseq_to_deseq2(ps, ~ Variety + Condition)
dds$Condition <- relevel(dds$Condition, "Control")  # Set Control as reference

dds <- DESeq(dds, test = "Wald", fitType = "parametric")

# Results for the average effect of inoculation (Inoculated vs Control)
res <- results(dds, contrast = c("Condition", "Inoculated", "Control"))

# Log2 Fold Change Shrinkage (standard for stable estimation)
# Ensure the coefficient matches resultsNames(dds)
res_shr <- lfcShrink(dds, coef = "Condition_Inoculated_vs_Control", type = "apeglm")

# 3) Merge Results with Taxonomy -------------------------------------------
out <- as.data.frame(res_shr) %>%
  mutate(pvalue = res$pvalue, 
         padj = res$padj,
         ASV = rownames(res_shr))

tax <- as.data.frame(as(tax_table(ps), "matrix"))
tax$ASV <- rownames(tax)

out <- left_join(out, tax, by = "ASV") %>%
  arrange(pvalue)

# 4) Significant ASV Summary -----------------------------------------------
sig <- out %>% filter(!is.na(pvalue), pvalue < 0.05)
message("Number of globally significant ASVs (p < 0.05): ", nrow(sig))

# Summary by Phylum
by_phylum <- sig %>%
  mutate(
    Phylum   = as.character(Phylum),
    enriched = as.integer(log2FoldChange > 0), # Enrichment indicator
    depleted = as.integer(log2FoldChange < 0)  # Depletion indicator
  ) %>%
  group_by(Phylum) %>%
  summarise(
    n_sig      = n(),
    n_enriched = sum(enriched, na.rm = TRUE),
    n_depleted = sum(depleted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_enriched))

print(by_phylum)

# 5) Volcano Plot ----------------------------------------------------------
ggplot(out, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = pvalue < 0.05), alpha = 0.6) +
  scale_color_manual(values = c("grey70", "#D55E00"), name = "Significant (p < 0.05)") +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_minimal() +
  labs(title = "Global Differential Abundance: Inoculated vs Control",
       subtitle = "Adjusted for Variety effect",
       x = "log2FoldChange (Inoculated / Control)", 
       y = expression(-log[10](p-value)))
#manhattan plot
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)
library(grid)

## ====== Parameters and Data Preparation ======
ps            <- ps16S_Root_10       
use_p         <- "pvalue"           # Options: "pvalue" or "padj"
alpha         <- 0.05
label_top     <- 25
jitter_w      <- 0.28
top_phyla     <- 10
size_breaks   <- c(0.01, 0.05, 0.10, 0.50, 1.00)

# Calculate Global Relative Abundance (%)
ra_pct <- 100 * taxa_sums(ps) / sum(taxa_sums(ps))
ra_df  <- tibble::tibble(ASV = names(ra_pct), rel_abund_pct = as.numeric(ra_pct))

# Format p-values and log-transformation
P_vec <- as.numeric(out[[use_p]])
P_vec[!is.finite(P_vec) | P_vec <= 0] <- NA_real_
y_vec <- -log10(pmin(pmax(P_vec, 1e-300), 1))

df0 <- out %>%
  dplyr::mutate(P = P_vec, y = y_vec) %>%
  dplyr::left_join(ra_df, by = "ASV") %>%
  dplyr::mutate(
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Others", as.character(Phylum)),
    Order  = ifelse(is.na(Order)  | Order  == "", "Unclassified", as.character(Order)),
    rel_abund_pct = tidyr::replace_na(rel_abund_pct, 0),
    sig = is.finite(P) & P < alpha
  ) %>%
  dplyr::filter(is.finite(y))

## ====== Manhattan Plot Function ======
make_manh <- function(df0, direction = c("enriched", "depleted"),
                      top_phyla = 12, alpha = 0.05, label_top = 25,
                      jitter_w = 0.28, bg_alpha = 0.10,
                      palette_cols = c("#78A1BB","#EEB902","#7776BC","#EC7357", "#28AFB0", 
                                       "#97CC04","#E13700","#C09BD8", "#337CA0")) {

  direction <- match.arg(direction)
  df_sub <- if (direction == "enriched") dplyr::filter(df0, log2FoldChange > 0)
            else                         dplyr::filter(df0, log2FoldChange < 0)

  # Order phyla by ASV count - Forced dplyr to avoid conflicts
  ph_order <- df_sub %>%
    dplyr::count(Phylum, name = "n") %>%
    dplyr::arrange(dplyr::desc(n)) %>%
    dplyr::pull(Phylum)

  keep <- head(ph_order, top_phyla)
  df_sub <- df_sub %>%
    dplyr::mutate(Phylum2 = ifelse(Phylum %in% keep, Phylum, "Others"))

  ph_axis <- c(keep, setdiff("Others", keep))

  # Index Orders within Phylum for X-axis positioning
  ord_map <- df_sub %>%
    dplyr::distinct(Phylum2, Order) %>%
    dplyr::arrange(factor(Phylum2, levels = ph_axis), Order) %>%
    dplyr::group_by(Phylum2) %>%
    dplyr::mutate(ord_idx = match(Order, unique(Order))) %>%
    dplyr::ungroup()

  # Calculate block positions
  ph_info <- ord_map %>%
    dplyr::group_by(Phylum2) %>%
    dplyr::summarise(n_ord = max(ord_idx), .groups = "drop") %>%
    dplyr::arrange(factor(Phylum2, levels = ph_axis)) %>%
    dplyr::mutate(start = c(0, head(cumsum(n_ord), -1)),
                  mid   = start + n_ord/2 + 0.5)

  # Coordinates and jitter
  df_plot <- df_sub %>%
    dplyr::left_join(ord_map, by = c("Phylum2", "Order")) %>%
    dplyr::left_join(ph_info, by = "Phylum2") %>%
    dplyr::mutate(x_core = start + ord_idx,
                  x      = x_core + runif(dplyr::n(), -jitter_w, jitter_w))

  # Color assignment
  n_needed  <- length(ph_axis)
  pal_auto  <- grDevices::colorRampPalette(palette_cols)(n_needed)
  pal       <- stats::setNames(pal_auto, ph_axis)
  pal["Others"] <- "#9E9E9E"

  # Background blocks
  bg <- ph_info %>%
    dplyr::mutate(xmin = start + 0.5, xmax = start + n_ord + 0.5)

  # Taxon labels
  lab_df <- df_plot %>%
    dplyr::filter(sig) %>%
    dplyr::arrange(P, dplyr::desc(abs(log2FoldChange))) %>%
    dplyr::slice_head(n = label_top) %>%
    dplyr::mutate(
      Genus   = na_if(trimws(as.character(Genus)), ""),
      Species = na_if(trimws(as.character(Species)), ""),
      Species = dplyr::if_else(
        !is.na(Species) & tolower(Species) %in% c("sp.", "sp", "uncultured", "uncultured bacterium"),
        NA_character_, Species
      ),
      line1 = dplyr::case_when(
        !is.na(Genus) & !is.na(Species) ~ paste0(Genus, "_", Species),
        !is.na(Genus)                   ~ paste0(Genus, "_sp."),
        TRUE                            ~ "Unassigned"
      ),
      line2 = paste0("(", ASV, ")"),
      label = paste0(line1, "\n", line2)
    )

  ttl <- if (direction == "enriched") "Global DA — Enriched ASVs (log2FC > 0)"
         else                         "Global DA — Depleted ASVs (log2FC < 0)"

  # Final ggplot
  p <- ggplot(df_plot, aes(x = x, y = y)) +
    geom_rect(data = bg,
              aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = Phylum2),
              inherit.aes = FALSE, alpha = bg_alpha, colour = NA, show.legend = FALSE) +
    geom_hline(yintercept = -log10(alpha), linetype = 2, color = "grey40") +
    geom_point(aes(color = Phylum2, size = rel_abund_pct), alpha = 0.9) +
    ggrepel::geom_text_repel(
      data = lab_df, aes(label = label, color = Phylum2),
      size = 3.1, max.overlaps = Inf, seed = 1,
      box.padding = 0.4, point.padding = 0.25, segment.alpha = 0.5, show.legend = FALSE
    ) +
    scale_color_manual(values = pal, name = "Phylum") +
    scale_fill_manual(values = pal, guide = "none") +
    guides(
      color = guide_legend(
        override.aes = list(shape = 15, size = 6, alpha = 1),  
        keywidth = unit(8, "mm"), keyheight = unit(8, "mm"),
        title.position = "top"
      ),
      size = guide_legend(order = 2) 
    ) +
    scale_size_continuous(
      range = c(1.6, 6.2),
      breaks = size_breaks, labels = paste0(size_breaks, "%"),
      name = "Relative abundance (%)"
    ) +
    scale_x_continuous(breaks = ph_info$mid, labels = ph_info$Phylum2, expand = c(0.01, 0.01)) +
    labs(
      title    = ttl,
      subtitle = paste0("ASVs plotted by -log10(", if (use_p=="padj") "FDR" else "p-value", ") vs Phylum"),
      x = NULL,
      y = if (use_p=="padj") expression(-log[10]("FDR")) else expression(-log[10](p-value))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "right",
      legend.key.size = unit(9, "mm"),
      legend.text  = element_text(size = 11),
      legend.title = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(angle = 25, hjust = 1),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )

  return(p)
}

## ====== Custom Palette ======
pal_user <- c("#78A1BB","#EEB902","#7776BC","#EC7357", "#28AFB0", 
              "#97CC04","#E13700","#C09BD8", "#337CA0")

## ====== Generate Plots ======
p_enr <- make_manh(df0, direction = "enriched",
                   top_phyla = top_phyla, alpha = alpha,
                   label_top = label_top, jitter_w = jitter_w,
                   palette_cols = pal_user)

p_dep <- make_manh(df0, direction = "depleted",
                   top_phyla = top_phyla, alpha = alpha,
                   label_top = label_top, jitter_w = jitter_w,
                   palette_cols = pal_user)

p_enr; p_dep

# Save plots
ggsave(file.path(output_dir_fig, paste0("Fig5_Manhattan_Enriched_", use_p, ".svg")),
       p_enr, width = 12, height = 7)

ggsave(file.path(output_dir_fig, paste0("Fig5_Manhattan_Depleted_", use_p, ".svg")),
       p_dep, width = 12, height = 7)
```

# --- FIG 6: ALDEX2 BY VARIETY ---
```{r}
library(ALDEx2)
library(tidyverse)
library(phyloseq)
library(openxlsx)

# Set seed for reproducibility (CRITICAL for ALDEx2)
set.seed(123)

varieties <- unique(as.character(sample_data(ps16S_Root_10)$Variety))
all_genus_results <- list()
all_asv_results <- list()

# ============================
# 📌 GENUS Loop
# ============================
for (variety in varieties) {
  message("🔹 Processing (GENUS) : ", variety)
  
  # Using your exact subsetting logic but secured for the loop
  ps_variety <- prune_samples(sample_data(ps16S_Root_10)$Variety == variety, ps16S_Root_10)
  
  ps_genus <- tax_glom(ps_variety, taxrank = "Genus", NArm = TRUE)
  
  otu <- otu_table(ps_genus)
  if (!taxa_are_rows(otu)) { otu <- t(otu) }
  otu_df <- as.data.frame(otu)
  
  condition <- as.character(sample_data(ps_genus)$Condition)
  
  # ALDEx2 process
  aldex_res <- aldex.clr(otu_df, conds = condition, mc.samples = 128, denom = "all", verbose = FALSE)
  aldex_tt  <- aldex.ttest(aldex_res, paired.test = FALSE)
  aldex_eff <- aldex.effect(aldex_res)
  
  res <- data.frame(aldex_tt, aldex_eff)
  res$Feature <- rownames(res)
  res$Variety <- variety
  
  tax <- as.data.frame(as(tax_table(ps_genus), "matrix"))
  tax$Feature <- rownames(tax)
  res_annot <- merge(res, tax, by = "Feature")
  
  # Your exact filter
  res_sig <- res_annot %>% filter(wi.ep < 0.05 & abs(diff.btw) > 1)
  
  all_genus_results[[variety]] <- res_sig
}

# ============================
# 📌 ASV Loop
# ============================
for (variety in varieties) {
  message("🔹 Processing (ASV) : ", variety)
  
  ps_variety <- prune_samples(sample_data(ps16S_Root_10)$Variety == variety, ps16S_Root_10)
  
  otu <- otu_table(ps_variety)
  if (!taxa_are_rows(otu)) { otu <- t(otu) }
  otu_df <- as.data.frame(otu)
  
  condition <- as.character(sample_data(ps_variety)$Condition)
  
  aldex_res <- aldex.clr(otu_df, conds = condition, mc.samples = 128, denom = "all", verbose = FALSE)
  aldex_tt  <- aldex.ttest(aldex_res, paired.test = FALSE)
  aldex_eff <- aldex.effect(aldex_res)
  
  res <- data.frame(aldex_tt, aldex_eff)
  res$ASV <- rownames(res)
  res$Variety <- variety
  
  tax <- as.data.frame(as(tax_table(ps_variety), "matrix"))
  tax$ASV <- rownames(tax)
  tax$Feature <- paste(tax$Genus, tax$Species, tax$ASV, sep = "_")
  res <- merge(res, tax, by = "ASV")
  
  # Your exact filter
  res_sig <- res %>% filter(wi.ep < 0.05 & abs(diff.btw) > 1)
  all_asv_results[[variety]] <- res_sig
}

# Merge and Export
final_genus_df <- bind_rows(all_genus_results)
final_asv_df   <- bind_rows(all_asv_results)



wb <- createWorkbook()
addWorksheet(wb, "Significant_Genus")
writeData(wb, "Significant_Genus", final_genus_df)
addWorksheet(wb, "Significant_ASVs")
writeData(wb, "Significant_ASVs", final_asv_df)
saveWorkbook(wb, output_dir_fig, overwrite = TRUE)

message("✅ Excel file successfully created: ", output_dir_fig)

library(dplyr)
library(ggplot2)
library(phyloseq)

# 1. Load ALDEx2 results (from Genus sheet)
aldex_plot_data <- final_genus_df

# 2. Calculate average relative abundance per Genus and Variety
# Transform counts to percentages using the non-rarefied filtered object
ps_rel <- transform_sample_counts(ps16S_Root_10, function(x) 100 * x / sum(x))
ps_genus_rel <- tax_glom(ps_rel, taxrank = "Genus", NArm = TRUE)

# Melt to data frame
melted_genus <- psmelt(ps_genus_rel)

# Group by Variety and Genus to get the mean
abundance_stats <- melted_genus %>%
  dplyr::group_by(Variety, Genus) %>%
  dplyr::summarise(MeanAbundance = mean(Abundance), .groups = "drop")

# 3. Merge ALDEx2 results with mean abundances
# Linking stats with real community proportions for point sizing
plot_df <- left_join(aldex_plot_data, abundance_stats, by = c("Variety", "Genus"))

# 4. Final Plot Preparation
plot_df <- plot_df %>%
  dplyr::filter(!is.na(diff.btw)) %>%
  # Reorder Genus by effect size for better visualization
  mutate(Genus = reorder(Genus, diff.btw))

# 5. Bubble Plot Visualization
p_aldex <- ggplot(plot_df, aes(x = diff.btw, y = Genus, color = Variety)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  
  # Points sized by mean relative abundance
  geom_point(aes(size = MeanAbundance), alpha = 0.8) +
  
  # Standardize colors based on your defined variety palette
  scale_color_manual(values = colors_Variety) + 
  
  # Point size scale
  scale_size_continuous(name = "Mean Rel. Abundance (%)", range = c(2, 8)) +
  
  # Labels and theme
  labs(
    title = "Differentially Abundant Genera (ALDEx2)",
    subtitle = "Point size represents mean relative abundance within each variety",
    x = "Median CLR Difference (Control < 0 < Inoculated)",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    
    # --- MODIFICATION 2 : Mettre le texte de l'axe Y en italique ---
    axis.text.y = element_text(face = "italic")
  )
print(p_aldex)

# Save Plot
ggsave(file.path(output_dir_fig, "Fig6_ALDEx2_Genus_BubblePlot.svg"), 
       plot = p_aldex, width = 6, height = 6)
```


# --- FIG 6: ALTERNATIVE DESEQ2 BY VARIETY ---
```{r}
library(DESeq2)
library(dplyr)
library(tidyr)

# 1. Setup ---------------------------------------------------------------------
# Use non-rarefied counts (ps16S_Root_10)
ps16S_Root_100 <- phyloseq::filter_taxa(ps16S_Root, function(x) sum(x) >= 100, prune = TRUE)

ps_ds_v <- ps16S_Root_100
variety_list <- unique(as.character(sample_data(ps_ds_v)$Variety))

# List to store significant results for each variety
deseq_variety_results <- list()

# 2. Iterative DESeq2 Analysis -------------------------------------------------
for (current_var in variety_list) {
  message("🔹 Processing DESeq2 for variety: ", current_var)
  
  # Subset phyloseq for the specific variety
  ps_sub <- prune_samples(sample_data(ps_ds_v)$Variety == current_var, ps_ds_v)
  
  # Ensure Condition is a factor with Control as reference
  sample_data(ps_sub)$Condition <- factor(sample_data(ps_sub)$Condition, levels = c("Control", "Inoculated"))
  
  # Convert to DESeq2 object
  # Design: ~ Condition (since we are looking at one variety at a time)
  dds_sub <- phyloseq_to_deseq2(ps_sub, ~ Condition)
  
  # Run DESeq2
  dds_sub <- DESeq(dds_sub, test = "Wald", fitType = "parametric")
  
  # Extract results (Inoculated vs Control)
  res_v <- results(dds_sub, contrast = c("Condition", "Inoculated", "Control"))
  
  # Merge with taxonomy
  res_df <- as.data.frame(res_v) %>%
    mutate(ASV = rownames(.), Variety = current_var)
  
  tax_df <- as.data.frame(as(tax_table(ps_sub), "matrix")) %>% rownames_to_column("ASV")
  res_df <- left_join(res_df, tax_df, by = "ASV")
  
  # Filter for nominally significant ASVs (unadjusted p-value < 0.05)
  res_sig <- res_df %>% 
    dplyr::filter(!is.na(pvalue), pvalue < 0.05)
  
  if(nrow(res_sig) > 0) {
    deseq_variety_results[[current_var]] <- res_sig
  }
}

# Combine all results into one dataframe
final_deseq_var_df <- bind_rows(deseq_variety_results)



# 1. Calculate Mean Relative Abundance -----------------------------------------
# We use percentages to scale the points in the bubble plot
ps_rel <- transform_sample_counts(ps_ds_v, function(x) 100 * x / sum(x))
ps_genus_rel <- tax_glom(ps_rel, taxrank = "Genus", NArm = TRUE)
melted_genus <- psmelt(ps_genus_rel)

abundance_stats <- melted_genus %>%
  dplyr::group_by(Variety, Genus) %>%
  dplyr::summarise(MeanAbundance = mean(Abundance), .groups = "drop")

# 2. Merge DESeq2 results with Abundance stats ---------------------------------
# We aggregate ASV results to Genus level for the plot
plot_deseq_df <- final_deseq_var_df %>%
  dplyr::group_by(Variety, Genus) %>%
  dplyr::summarise(log2FoldChange = mean(log2FoldChange), # Mean effect size per genus
                   pvalue = min(pvalue),                 # Best p-value per genus
                   .groups = "drop") %>%
  left_join(abundance_stats, by = c("Variety", "Genus"))

# 3. Final Plotting ------------------------------------------------------------
plot_deseq_df <- plot_deseq_df %>%
  dplyr::filter(!is.na(Genus)) %>%
  mutate(Genus = reorder(Genus, log2FoldChange))

p_deseq_var <- ggplot(plot_deseq_df, aes(x = log2FoldChange, y = Genus, color = Variety)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(size = MeanAbundance), alpha = 0.7) +
  
  # Use your defined variety colors
  scale_color_manual(values = colors_Variety) + 
  scale_size_continuous(name = "Mean Rel. Abundance (%)", range = c(2, 8)) +
  
  labs(
    title = "Differentially Abundant Genera per Variety (DESeq2)",
    subtitle = "Points represent nominal significance (unadjusted p < 0.05)",
    x = "log2FoldChange (Inoculated / Control)",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  )

print(p_deseq_var)

# Save the plot
ggsave(file.path(output_dir_fig, "Fig6_DESeq2_Variety_Specific_BubblePlot.svg"), 
       plot = p_deseq_var, width = 10, height = 7)
```


#Supplementary figures
# --- FIG S1: RAREFACTION CURVES ---
```{r}
RCall <- ggrare(ps16S_Root, step = 100, se = FALSE, color = "Variety") + 
  labs(title = "Rarefaction curve 16S, depending on the variety (no_norm)", 
       x = "Depth",
       y = "Number of ASVs") +
  theme_minimal(base_size = 15) + 
    scale_color_manual(values = c("Leak" = "#7CAE00", "PhkarRumduol" = "#00BFC4", "PhkarMalis"="#FF3B0A","Russey"="#FFC000", "IR64"="#9E1FFF")) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)) +
  coord_cartesian(xlim = c(0, 22000))
print(RCall)

# Saving..
ggsave(filename = file.path(output_dir_fig, "Rarefaction_curve_16S_variety_Root_noPGM_nonorm.jpg"),
       plot = RCall, width = 10, height = 7, dpi = 300)



ps16S_Root_rare <- rarefy_even_depth(ps16S_Root, rngseed = 123,  sample.size = min(sample_sums(ps16S_Root)),replace = FALSE, verbose = TRUE)
ps16S_Root_rare

RCall <- ggrare(ps16S_Root_rare, step = 100, se = FALSE, color = "Variety") + 
  labs(title = "Rarefaction curve 16S, depending on the variety (rare)", 
       x = "Depth",
       y = "Number of ASVs") +
  theme_minimal(base_size = 15) + 
    scale_color_manual(values = c("Leak" = "#7CAE00", "PhkarRumduol" = "#00BFC4", "PhkarMalis"="#FF3B0A","Russey"="#FFC000", "IR64"="#9E1FFF")) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)) +
  coord_cartesian(xlim = c(0, 17000))
print(RCall)

# Enregistrer le graphique
ggsave(filename = file.path(output_dir_fig, "FigS1_Rarefaction_curve_16S_variety_Root_noPGM_rare.jpg"),
       plot = RCall, width = 10, height = 7, dpi = 300)
```





# --- FIG S4: PHENOTYPE ---
```{r}
# ==============================================================================
# PHENOTYPIC DATA ANALYSIS
# ==============================================================================

library(dplyr)
library(ggplot2)
library(ggpubr)
library(multcompView) # Required for significance letters

# 1. Data Extraction & Setup ---------------------------------------------------
# Extract metadata from the phyloseq object to ensure consistency
pheno_df <- data.frame(sample_data(ps16S_Root_10_rare))

# Ensure columns are numeric
pheno_df$Symptoms_up <- as.numeric(pheno_df$Symptoms_up)
pheno_df$Size <- as.numeric(pheno_df$Size)

# Define Palette
colors_Variety <- c("Leak" = "#7CAE00", "PhkarRumduol" = "#00BFC4", 
                    "PhkarMalis" = "#FF3B0A", "Russey" = "#FFC000", "IR64" = "#9E1FFF")

# ------------------------------------------------------------------------------
# FUNCTION: Generate Significance Letters (Kruskal + Pairwise Wilcoxon)
# ------------------------------------------------------------------------------
get_significance_letters <- function(data, value_col, group_col) {
  # 1. Pairwise Wilcoxon Test (Non-parametric) with BH adjustment
  pwc <- pairwise.wilcox.test(data[[value_col]], data[[group_col]],
                              p.adjust.method = "none")
  
  # 2. Convert p-values to a matrix for multcompView
  p_mat <- pwc$p.value
  
  # Handles cases where the matrix is not square (required for multcompLetters)
  # We convert the triangular output to a full matrix/list format if needed
  # Using a helper logic for multcompLetters:
  
  # Format p-values into a named vector for multcompLetters
  # Create a logical vector: TRUE if p < 0.05
  # Note: multcompLetters is tricky with pairwise.wilcox output.
  # We will use a robust method: compare everything to everything manually for the input string.
  
  # Alternative: Use "tri.to.squ" logic or simply loop.
  # Let's use the matrix directly if compatible, or reconstruct names.
  
  # Simplified approach using multcompLetters directly on p-matrix usually requires full matrix.
  # To avoid errors, we calculate letters based on the p-value matrix:
  
  # If p_mat contains NA, replace with 1 (self-comparison) or handle.
  # Actually, let's use a try-catch standard loop to generate the comparison vector.
  
  # Convert triangular p-matrix to a list of comparisons "A-B" = p.val
  p_vec <- c()
  rn <- rownames(p_mat)
  cn <- colnames(p_mat)
  
  for(r in rn) {
    for(c in cn) {
      if(!is.na(p_mat[r, c])) {
        comp_name <- paste(r, c, sep = "-")
        p_vec[comp_name] <- p_mat[r, c]
      }
    }
  }
  
  # Generate letters
  letters_list <- multcompLetters(p_vec, compare = "<", threshold = 0.05, Letters = letters)
  
  # Create a dataframe for plotting
  res_df <- data.frame(
    Variety = names(letters_list$Letters),
    Letter = letters_list$Letters
  )
  return(res_df)
}

# ==============================================================================
# FIGURE A: SYMPTOMS (Inoculated Plants Only)
# ==============================================================================

# 1. Subset Data
df_symp <- pheno_df %>% 
  filter(Condition == "Inoculated") %>%
  filter(!is.na(Symptoms_up))

# 2. Calculate Statistics (Letters)
letters_symp <- get_significance_letters(df_symp, "Symptoms_up", "Variety")

# 3. Calculate Y position for letters (Max value + 10% buffer)
max_vals_symp <- df_symp %>% 
  group_by(Variety) %>% 
  summarise(MaxY = max(Symptoms_up, na.rm=TRUE)) %>%
  left_join(letters_symp, by = "Variety")

# 4. Plot
p_symptoms <- ggplot(df_symp, aes(x = Variety, y = Symptoms_up, fill = Variety)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  
  # Add Letters
  geom_text(data = max_vals_symp, aes(x = Variety, y = MaxY, label = Letter), 
            vjust = -0.5, size = 5, fontface = "bold", inherit.aes = FALSE) +
  
  scale_fill_manual(values = colors_Variety) +
  labs(title = "Symptom Severity (Inoculated Plants)",
       y = "Lesion Length (cm)",
       x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Add margin at top for letters
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

print(p_symptoms)
ggsave(file.path(output_dir_fig, "FigS4_Boxplot_Symptoms_Inoculated.svg"), p_symptoms, width = 6, height = 4)

# ==============================================================================
# FIGURE B: PLANT SIZE (Control/Healthy Plants Only)
# ==============================================================================

# 1. Subset Data
df_size <- pheno_df %>% 
  filter(Condition == "Control") %>%
  filter(!is.na(Size))

# 2. Calculate Statistics (Letters)
letters_size <- get_significance_letters(df_size, "Size", "Variety")

# 3. Calculate Y position for letters
max_vals_size <- df_size %>% 
  group_by(Variety) %>% 
  summarise(MaxY = max(Size, na.rm=TRUE)) %>%
  left_join(letters_size, by = "Variety")

# 4. Plot
p_size <- ggplot(df_size, aes(x = Variety, y = Size, fill = Variety)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
  
  # Add Letters
  geom_text(data = max_vals_size, aes(x = Variety, y = MaxY, label = Letter), 
            vjust = -0.5, size = 5, fontface = "bold", inherit.aes = FALSE) +
  
  scale_fill_manual(values = colors_Variety) +
  labs(title = "Plant Growth (Healthy/Control Plants)",
       y = "Plant Height (cm)",
       x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

print(p_size)
ggsave(file.path(output_dir_fig, "FigS4_Boxplot_Size_Control.svg"), p_size, width = 6, height = 4)
```

#PERMANOVA
```{r}
library(vegan)

# 1. Prepare the distance matrix (Bray-Curtis)
# Ensure the OTU table is correctly oriented (samples in rows)
otu_table_matrix <- as.matrix(otu_table(ps16S_Root_10_rare))
bray_dist <- vegdist(otu_table_matrix, method = "bray")

# 2. Extract metadata
sample_metadata <- data.frame(sample_data(ps16S_Root_10_rare))

# 3. Global PERMANOVA: Testing the effects of Variety, Condition, and their Interaction
# This is usually the main table presented in a publication
permanova_global <- adonis2(
  bray_dist ~ Variety * Condition,
  data = sample_metadata,
  permutations = 999,
  method = "bray")

# Print results to console
print("Global PERMANOVA Results:")
print(permanova_global)

# Export Global PERMANOVA results
write.csv(as.data.frame(permanova_global), 
          file = file.path(output_dir_fig, "Table1_Global_PERMANOVA_Variety_Condition.csv"))
```
```{r}
# Note: Ensure symptoms data is merged in the metadata as done in the previous sections
# Testing if Symptoms_up significantly explains microbiota variance across varieties
permanova_symptoms <- adonis2(
  bray_dist ~ Symptoms_up + Variety,
  data = sample_metadata, # Using the same metadata containing symptoms
  permutations = 999,
  method = "bray",
  na.action = na.omit # Handles samples without symptom data
)

print("PERMANOVA Results (Symptoms + Variety):")
print(permanova_symptoms)

# Export results
write.csv(as.data.frame(permanova_symptoms), 
          file = file.path(output_dir_fig, "Table1_PERMANOVA_Symptoms_Variety.csv"))
```

# END